---
title: "Managing and Understanding Data"
author: "Escribir vuestro nombre y apellidos"
date: '`r format(Sys.Date(),"%e de %B, %Y")`'
output:
  html_document:
    toc: yes
  pdf_document:
    toc: yes
bibliography: scholar.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load_libraries, include=FALSE}
library(knitr)
```
# R data structures

The R data structures used most frequently in machine learning are *vectors*, *factors*, *lists*, *arrays*,
and *data frames*.

To find out more about statistical methods with R see [@teetor2011r;@hothorn2014handbook;@baayen2008analyzing].

## Vectors 

The fundamental R data structure is the **vector**, which stores an ordered set of
values called **elements**. A vector can contain any number of elements. However, all
the elements must be of the same type; for instance, a vector cannot contain both
numbers and text.

There are several vector types commonly used in machine learning:`integer`(numbers without decimals), `numeric` (numbers with decimals), `character` (text data),
or `logical` (`TRUE` or `FALSE` values). There are also two special values: `NULL`, which is
used to indicate the absence of any value, and `NA`, which indicates a missing value.

Create vectors of data for three medical patients:

```{r chunck1}
# create vectors of data for three medical patients
subject_name <- c("John Doe", "Jane Doe", "Steve Graves")
temperature <- c(98.1, 98.6, 101.4)
flu_status <- c(FALSE, FALSE, TRUE)
```

Access the second element in body temperature vector:

```{r chunck2}
# access the second element in body temperature vector
temperature[2]
```

Examples of accessing items in vector include items in the range 2 to 3.

```{r chunck3}
## examples of accessing items in vector
# include items in the range 2 to 3
temperature[2:3]
```

Exclude item 2 using the minus sign
```{r chunck4}
# exclude item 2 using the minus sign
temperature[-2]

```

Use a vector to indicate whether to include item

```{r chunk 5}
# use a vector to indicate whether to include item
temperature[c(TRUE, TRUE, FALSE)]

```

# Exploring and understanding data

After collecting data and loading it into R data structures, the next step in the
machine learning process involves examining the data in detail. It is during this
step that you will begin to explore the data's features and examples, and realize the
peculiarities that make your data unique. The better you understand your data, the
better you will be able to match a machine learning model to your learning problem.
The best way to understand the process of data exploration is by example. In this
section, we will explore the `usedcars.csv` dataset, which contains actual data about
used cars recently advertised for sale on a popular U.S. website.

Since the dataset is stored in CSV form, we can use the `read.csv()` function to load
the data into an R data frame:

```{r read}
##### Exploring and understanding data --------------------

## data exploration example using used car data
usedcars <- read.csv("usedcars.csv", stringsAsFactors = FALSE)

```

## Exploring the structure of data
One of the first questions to ask in your investigation should be about how data
is organized. If you are fortunate, your source will provide a **data dictionary**, a
document that describes the data's features. In our case, the used car data does not
come with this documentation, so we'll need to create our own.


```{r str}
# get structure of used car data
str(usedcars)

```

## Exploring numeric variables


```{r explore}
## Exploring numeric variables -----

# summarize numeric variables
summary(usedcars$year)
summary(usedcars[c("price", "mileage")])

# calculate the mean income
(36000 + 44000 + 56000) / 3
mean(c(36000, 44000, 56000))

# the median income
median(c(36000, 44000, 56000))

# the min/max of used car prices
range(usedcars$price)

# the difference of the range
diff(range(usedcars$price))

# IQR for used car prices
IQR(usedcars$price)

# use quantile to calculate five-number summary
quantile(usedcars$price)

# the 99th percentile
quantile(usedcars$price, probs = c(0.01, 0.99))

# quintiles
quantile(usedcars$price, seq(from = 0, to = 1, by = 0.20))
```


## Visualizing numeric variables - boxplots

```{r graphics, fig.height=4, fig.width=6}

# boxplot of used car prices and mileage
boxplot(usedcars$price, main="Boxplot of Used Car Prices",ylab="Price ($)")

boxplot(usedcars$mileage, main="Boxplot of Used Car Mileage",
      ylab="Odometer (mi.)")

# histograms of used car prices and mileage
hist(usedcars$price, main = "Histogram of Used Car Prices",
     xlab = "Price ($)")

hist(usedcars$mileage, main = "Histogram of Used Car Mileage",
     xlab = "Odometer (mi.)")

# variance and standard deviation of the used car data
var(usedcars$price)
sd(usedcars$price)
var(usedcars$mileage)
sd(usedcars$mileage)

```

## Measuring spread - quartiles and the five-number summary

The **five-number summary** is a set of five statistics that roughly depict the spread
of a dataset. All five of the statistics are included in the output of the `summary()`
function. Written in order, they are:

1. Minimum (`Min.`)
2. First quartile, or Q1 (`1st Qu.`)
3. Median, or Q2 (`Median`)
4. Third quartile, or Q3 (`3rd Qu.`)
5. Maximum (`Max.`)

## Measuring spread - variance and standard deviation

In order to calculate the standard deviation, we must first obtain the **variance**, which
is defined as the average of the squared differences between each value and the
mean value. In mathematical notation, the variance of a set of `n` values of `x` is defined
by the following formula. The Greek letter `mu` ($\mu$) (similar in appearance to an m)
denotes the mean of the values, and the variance itself is denoted by the Greek letter
`sigma` ($\sigma$) squared (similar to a b turned sideways):

$$
Var(X)= \sigma^2 = \frac{1}{n}\sum_{i=1}^n (x_i - \mu)^2
$$


The standard deviation is the square root of the variance, and is denoted by `sigma` as
shown in the following formula:

$$
StdDev(X)= \sigma = \sqrt{\frac{1}{n}\sum_{i=1}^n (x_i - \mu)^2}
$$


*Note*. For more details on using mathematical expressions in Latex (R Markdown) see <https://es.sharelatex.com/learn/Mathematical_expressions>.



##Table with information about mileage and price


```{r}
mileage_s<-summary(usedcars$mileage)
price_s<-summary(usedcars$price)
kable(rbind(mileage_s,price_s))
```

##Some descriptive graphics

```{r}
par(mfrow=c(2,2))
hist(usedcars$mileage,xlab="Mileage",main="Histogram of mileage",col="grey85")
hist(usedcars$price,xlab="Price",main="Histogram of price",col="grey85")
usedcars$transmission<-factor(usedcars$transmission)
plot(usedcars$mileage,usedcars$price,pch=16,col=usedcars$transmission,xlab="Mileage",ylab="Price")
legend("topright",pch=16,c("AUTO","MANUAL"),col=1:2,cex=0.5)
```

#References

